---
title: "Homework 5"
format:
  pdf:
    fontsize: 12pt
    geometry:
      - inner=1.5cm
      - outer=1.5cm
      - top=2.5cm
      - bottom=2.5cm
    include-in-header:
      - text: |
          \addtokomafont{disposition}{\rmfamily}
          \usepackage{fancyhdr, lastpage, framed, caption, xcolor}
          \captionsetup[figure]{labelformat=empty}
          \pagestyle{fancyplain}
          \fancyhf{}
          \lhead{\fancyplain{}{STAT 0219: Time Series Analysis}}
          \rhead{\fancyplain{}{Stratton - HW4}}
          \fancyfoot[RO, LE] {page \thepage\ of \pageref{LastPage}}
          \thispagestyle{plain}
editor: source
---
<!-- Note: the YAML header is everything above this line. -->
\vspace{-3in}
\textbf{Name:} Your name here  
\textbf{Due:} 2024/10/14
\vspace{1.5in}

Be sure to submit **both** the .pdf and .qmd file to Canvas by Monday, October 14th at 11:59 pm. 

```{r, include = F}
set.seed(10052024)
rm(list = ls())
library(tidyverse)
```

0. [1 pt] With whom did you work on this assignment?

   :::{.callout-warning icon=false appearance="simple"}
   
   :::

1. [12 pt] The purpose of this question is to demonstrate that there exist time series structures that Holt-Winters cannot estimate very well. To demonstrate this idea, we are going to simulate data from an AR(2) process with $\alpha_1 = 1.1$ and $\alpha_2 = -.3$. 

   a) [1 pt] Write the hypothetical AR(2) model in proper notation.
   
      :::{.callout-warning icon=false appearance="simple"}

      :::   
      
   b) [1 pt] Express the model from part a in terms of the backshift operator. 
   
      :::{.callout-warning icon=false appearance="simple"}

      ::: 
      
   c) [1 pt] Determine whether the model is stationary. 
   
      :::{.callout-warning icon=false appearance="simple"}

      ::: 
      
```{r, indent = "      "}

```

   d) [3 pt] Simulate 12 years of monthly data from an AR(2) process with $\alpha_1 = 1.1$ and $\alpha_2 = -.3$. Convert the synthetic data to a `ts` object and plot the resulting series. 

```{r, indent = "      ", fig.dim = c(6, 4)}

```

   e) [1 pt] Create an ACF plot of the synthetic series and comment what the plot suggests about the autocorrelation.  
   
      :::{.callout-warning icon=false appearance="simple"}

      ::: 

```{r, indent = "      ", fig.dim = c(6, 3.25)}

```

   f) [1 pt] Create an PACF plot of the synthetic series and comment on what the plot suggests about the autocorrelation.  
   
      :::{.callout-warning icon=false appearance="simple"}

      ::: 

```{r, indent = "      ", fig.dim = c(6, 3.25)}

```

   g) [2 pt] Fit a Holt-Winters model to the synthetic series and plot the results. 
  
```{r, indent = "      ", fig.dim = c(6, 4)}

```

   h) [2 pt] Create an ACF plot of the residuals from the Holt-Winters model and comment on whether the model has accounted for the residual autocorrelation. 
   
      :::{.callout-warning icon=false appearance="simple"}

      ::: 

```{r, indent = "      ", fig.dim = c(6, 4)}

```

\newpage

2. [2 pt] Using the characteristic equation, show that a random walk is not stationary. 

   :::{.callout-warning icon=false appearance="simple"}

   :::

3. [3 pt] Determine for what values of $\alpha$ an AR(1) process is stationary. 

   :::{.callout-warning icon=false appearance="simple"}

   :::

4. [6 pt] The purpose of this question is to prove one of the second order properties of an AR(1) process. 

   a) [1 pt] Express an AR(1) process in terms of the backshift operator.  

      :::{.callout-warning icon=false appearance="simple"}

      :::
      
   b) [3 pt] Using the expression from part a, show that $x_t = \sum_{i=0}^\infty \alpha^i w_{t-i}$. It is likely helpful to know that $(1-B)^{-1} = 1 + B + B^2 + \dots$ by Binomial expansion.

      :::{.callout-warning icon=false appearance="simple"}

      :::
      
   c) [2 pt] Using the expression from part b, show that the $E(x_t) = 0$. Recall that $E\left( \sum_{i=1}^n X_i \right) = \sum_{i=1}^n E(X_i)$ and that $E(cX) = cE(X)$ if $c$ is a constant. 

      :::{.callout-warning icon=false appearance="simple"}

      :::

