---
title: "Exam 2"
format:
  pdf:
    fontsize: 12pt
    geometry:
      - inner=1.5cm
      - outer=1.5cm
      - top=2.5cm
      - bottom=2.5cm
    include-in-header:
      - text: |
          \addtokomafont{disposition}{\rmfamily}
          \usepackage{fancyhdr, lastpage, framed, caption, xcolor, setspace}
          \captionsetup[figure]{labelformat=empty}
          \pagestyle{fancyplain}
          \fancyhf{}
          \lhead{\fancyplain{}{STAT 0219: Time Series Analysis}}
          \rhead{\fancyplain{}{Stratton - Exam 2}}
          \fancyfoot[RO, LE] {page \thepage\ of \pageref{LastPage}}
          \thispagestyle{plain}
editor: source
---
<!-- Note: the YAML header is everything above this line. -->
\vspace{-3in}
\textbf{Name:}   
\textbf{Due:} 2024/12/01
\vspace{1.5in}

Be sure to submit **both** the .pdf and .qmd file to Canvas by Sunday, December 1st at 11:59 pm. The purpose of this exam is to synthesize some of the concepts we have discussed this semester, particularly with respect to forecasting and inference. 

```{r, message = F, warning = F, echo = T, include = F}
rm(list = ls())
# library packages for data manipulation
library(tidyverse)
set.seed(11142024)
```

:::{.callout-important}
You may use any resource to complete this except **except** for other people (classmates, other faculty members, your parents, etc.) Please do not discuss this exam with anyone but me, and be sure to **cite all references and materials used to answer each question.** Please type your name below to acknowledge that you have not discussed this exam with anyone else. 
:::

Your name:

\vspace{1in}

\newpage

**Question 1A [27 pt]** 

```{r, include = F, eval = F}
# leaving this in the exam so that you can see what was done to clean the data
# read in the data
fish <- read_csv("fish/TimeSeriesLogMeans.csv")

# create some date identifiers, impute missing values,
# and generally clean things up
# you are welcome :)
fish_cleanish <- fish %>%
  # lower case column names
  rename_with(tolower) %>% 
  # remove variance columns - not needed
  dplyr::select(-va1cover, -va2loghetads, -va4loggambo) %>%
  # create dates, roughly assuming dateseq = month
  group_by(location) %>%
  mutate(
    year = rep(1999:2016, each = 12)[9:(n()+8)],
    month = rep(1:12, 12)[9:(n()+8)] %>% factor(),
    dt = ym(paste0(year, "-", month))
  ) %>%
  ungroup() %>%
  mutate(row_id = 1:n())

# impute some missing values
library(imputeTS)
out <- list()
for(group in c("TP", "WRHW", "WRGP")){
  tmp <- filter(fish_cleanish, location == group)
  hetad_ts <- ts(
    tmp$me2loghetads,
    start = c(1999, 9),
    freq = 12
  )
  gambo_ts <- ts(
    tmp$me4loggambo,
    start = c(1999, 9),
    freq = 12
  )
  tmp$me2loghetads <- c(imputeTS::na_kalman(hetad_ts, model = "auto.arima"))
  tmp$me4loggambo <- c(imputeTS::na_kalman(gambo_ts, model = "auto.arima"))
  out[[group]] <- tmp
}
fish_clean <- do.call("rbind", out)
saveRDS(fish_clean, "fish_clean.rds")
```

Understanding the dynamics between predator species, prey species, and how each group leverages their environment is a fundamental objective of ecology. To explore these dynamics, researchers collected monthly measurements of the activity of a predatory fish, the Eastern Mosquitofish (*Gambusia holbrooki*) and its littoral zone prey species, the least Killifish (*Heteraandria formosa*) in three regions of Northern Florida (`location`) using throw traps.[^1] At each survey event, a throw net was cast three times and the average log density of the Mosquitofish (`me4loggambo`) and Killifish (`me2loghetads`) was recorded. Researchers also collected the average percent vegetative cover within the survey location (`cover1`), which is a measure of how the Killifish utilize surrounding vegetation to escape the Mosquitofish. Note that while no trends in fish activity over time are expected, researchers do anticipate monthly seasonal effects resulting from changes in temperature throughout the year.

\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{killifish}
  \captionof{figure}{Least Killifish}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=.7\linewidth]{mosquito}
  \captionof{figure}{Eastern Mosquitofish}
\end{minipage}
\end{figure}

**Researchers are interested in determining whether the relationship between the Killifish density and the percent vegetative cover depends upon the location, after accounting for the density of the Mosquitofish and any seasonal effects due to month.** Such a result would imply that the Killifish utilize the vegetative cover differently across the three regions, even after accounting for seasonal differences and the impact of the predator species. Within the .qmd file, there is a code chunk that cleans the data for you; you should use the `fish_clean.rds` file for this analysis. For this first question, your goal is to use **ordinary least squares** to address the research question. Be sure to address the model assumptions, but you may ignore any violations of the independence assumption when answering the research question. Your response should be formatted as a paragraph, which references an exploratory graphic, the model assumptions, and the results of the statistical test. Use the following rubric to guide your response:

- [3 pt] Exploratory visual that addresses the research question is created and discussed
- [6 pt] Appropriate statistical model(s) used
- [6 pt] Model assumptions appropriately addressed
- [4 pt] Appropriate statistical test used to address research question
- [4 pt] A conclusion, supported by statistical evidence, is given and the evidence is referenced (statistic and p-value)
- [4 pt] Formatting (written response, output is reasonably clean, no callout box errors, complete sentences, spelling, figures of reasonable size, etc.)


[^1]: A [link](https://datadryad.org/stash/dataset/doi:10.5061/dryad.qjq2bvqpw#methods) to the description of the data if you are curious. You do not need to download anything from this link. 

\newpage

```{r, message = F, fig.dim = c(7, 6)}
# read data
fish_clean <- readRDS("fish_clean.rds")

# exploratory plots

# statistical models

# plots for assumptions

# tests

```

:::{.callout-warning icon=false appearance="simple"}

::: 

\newpage

**Question 1B [24 pt]**

This is a continuation of question 1A. For this question, your goal is to use **generalized least squares** to address the research question. Be sure to address the model assumptions, and show that the model you fit resolves any violations of the independence assumption. Your response should be formatted as a paragraph, which references the model assumptions, proof that the model you fit addresses any serial correlation in the error structure, and the results of the statistical test. Use the following rubric to guide your response:

- [6 pt] Appropriate statistical model(s) used
- [6 pt] Model assumptions appropriately addressed. You need only reassess the independence assumption.
- [4 pt] Appropriate statistical test used to address research question
- [4 pt] A conclusion, supported by statistical evidence, is given and the evidence is referenced (statistic and p-value)
- [4 pt] Formatting (written response, output is reasonably clean, no callout box errors, complete sentences, spelling, figures of reasonable size, etc.)

\newpage

```{r, message = F, fig.dim = c(7, 6)}
# models

# assumptions

# tests

```

:::{.callout-warning icon=false appearance="simple"}

::: 

\newpage

**Question 2 [17 pt]**

For this question, we return to the Mauna Loa CO$_2$ measurements. As a reminder, this data set describes the average monthly CO$_2$ readings from the Mauna Loa observatory[^2], which have been collected continuously since March of 1958. Suppose that researchers are interested in determining when the CO$_2$ emissions from Mauna Loa are expected to exceed 450 parts per million.

![Source: [USGS](https://www.cnn.com/2022/11/29/us/mauna-loa-volcano-eruption-affect-climate-tool/index.html)](ml.jpg){width=80%}



**Your goal is to fit a model to the Mauna Loa time series, use that model to forecast the CO$_2$ emissions, and determine the month during which you are 95% confident that the CO$_2$ emissions are at least 450 ppm.** Your response should be formatted as a paragraph, and include the following:

- [3 pt] Define the model (what kind of model is fit and whether it contains regression coefficients)
- [4 pt] Proof that this model accounts for the serial correlation in the errors
- [4 pt] A plot that visualizes the observed series, fitted series, and forecasted series (with 95% prediction intervals) in a single figure
- [4 pt] An answer to the research question
- [2 pt] Good formatting (written response, output is reasonably clean, no callout box errors, complete sentences, spelling, figures of reasonable size, etc.)

[^2]: [Link](https://gml.noaa.gov/ccgg/trends/data.html) to the NOAA webpage containing the data. You do not need to download anything from this website. You should use the `co2_mm_mlo.csv` file.  

\newpage

```{r, message = F, warning = F, include = F}
# clean data
ml <- read_csv("co2_mm_mlo.csv", skip = 40)
ml_clean <- ml %>%
  dplyr::select(year, month, average) %>%
  mutate(dt = ym(paste0(year, "-", month)))
```

```{r, fig.dim = c(7, 6), message = F}

```


