---
title: "Homework 8"
format:
  pdf:
    fontsize: 12pt
    geometry:
      - inner=1.5cm
      - outer=1.5cm
      - top=2.5cm
      - bottom=2.5cm
    include-in-header:
      - text: |
          \addtokomafont{disposition}{\rmfamily}
          \usepackage{fancyhdr, lastpage, framed, caption, xcolor}
          \captionsetup[figure]{labelformat=empty}
          \pagestyle{fancyplain}
          \fancyhf{}
          \lhead{\fancyplain{}{STAT 0219: Time Series Analysis}}
          \rhead{\fancyplain{}{Stratton - HW8}}
          \fancyfoot[RO, LE] {page \thepage\ of \pageref{LastPage}}
          \thispagestyle{plain}
editor: source
---
<!-- Note: the YAML header is everything above this line. -->
\vspace{-3in}
\textbf{Name:} Your name here  
\textbf{Due:} 2024/11/18
\vspace{1.5in}

```{r, include = F}
rm(list= ls())
library(tidyverse)
library(nlme)
```

Be sure to submit **both** the .pdf and .qmd file to Canvas by Monday, November 18th at 11:59 pm. The purpose of this homework is to ensure that you are making meaningful progress on your project prior to the exam and break. 

**THE NORMAL LATE POLICY DOES NOT APPLY FOR THIS ASSIGNMENT. LATE WORK WILL NOT BE ACCEPTED AND WILL RECEIVE A 0.**

1. [1 pt] List all group members for your project. You can all turn in the same copy of this assignment. 

   :::{.callout-warning appearance="simple" icon=false}
   
   :::
   
2. [1 pt] Provide a link to the data set that you are using for the project. If the data are not found online, describe where you obtained the data.

   :::{.callout-warning appearance="simple" icon=false}
   
   :::   
   
3. [2 pt] Define the research question for your project.

   :::{.callout-warning appearance="simple" icon=false}
   
   :::   
   
4. [3 pt] Does your research question entail inference, forecasting, or both?

   :::{.callout-warning appearance="simple" icon=false}
   
   :::  
   
5. [4 pt] What statistical method will you use to address your research question? Give me a little meat here. For example:

   - Will you use purely stochastic approaches (i.e. `auto.arima` without regressors) or regression?
   - If forecasting, how will you assess the quality of your model?
   - If conducting inference, to what models will you compare to answer your research question?

   :::{.callout-warning appearance="simple" icon=false}
   
   :::  

6. [5 pt] Read your data set into R and create a figure that visualizes your research question. Comment on what the figure suggests about your research question (you do not need to fit any models yet - save that for the project). 

```{r, indent = "   ", fig.dim = c(7, 3.5)}

```

   :::{.callout-warning appearance="simple" icon=false}
   
   :::

