---
title: "Exam 1"
format:
  pdf:
    fontsize: 12pt
    geometry:
      - inner=1.5cm
      - outer=1.5cm
      - top=2.5cm
      - bottom=2.5cm
    include-in-header:
      - text: |
          \addtokomafont{disposition}{\rmfamily}
          \usepackage{fancyhdr, lastpage, framed, caption, xcolor}
          \captionsetup[figure]{labelformat=empty}
          \pagestyle{fancyplain}
          \fancyhf{}
          \lhead{\fancyplain{}{STAT 0219: Time Series Analysis}}
          \rhead{\fancyplain{}{Stratton - Exam 1}}
          \fancyfoot[RO, LE] {page \thepage\ of \pageref{LastPage}}
          \thispagestyle{plain}
editor: source
---
<!-- Note: the YAML header is everything above this line. -->
\vspace{-3in}
\textbf{Name:}   
\textbf{Due:} 2024/10/21
\vspace{1.5in}

Be sure to submit **both** the .pdf and .qmd file to Canvas by Friday, October 25th at 11:59 pm. 

```{r, message = F, warning = F, echo = T, include = F}
rm(list = ls())
# library packages for data manipulation
library(tidyverse)
set.seed(10122024)
```

:::{.callout-important}
You may use any resource to complete this except **except** for other people (classmates, other faculty members, your parents, etc.) Please do not discuss this exam with anyone but me, and be sure to **cite all references and materials used to answer each question.** Please type your name below to acknowledge that you have not discussed this exam with anyone else. 
:::

Your name:

\newpage

**Question 1 [42 pt]** 

The Utah Frontier Observatory for Research in Geothermal Energy (FORGE) is a project designed to collect and disseminate data that enables research in the development of geothermal energy. In February of 2021, the UTAH-FORGE drilled a 9000 foot hole, labeled well 56-32, to enable future monitoring of seismic activity. The `utah_forge_56_32_2021_02_08_03_04.csv` file contains a subset of the measurements obtained from the drilling process at 10 second intervals between 3 and 4 hours into the drilling procedure. The entire dataset is publicly available on [data.gov](https://catalog.data.gov/dataset/utah-forge-well-56-32-drilling-data-and-logs) and a description of the drilling site is provided on the Utah FORGE [website](https://utahforge.com/2021/03/25/drilling-of-well-56-32/). For this analysis, we are interested in the total volume of mud lost to drilling at each time point, labeled `Total Mud Volume` in the data set. 

![Drilling site, courtesy of [Utah FORGE](https://utahforge.com/2021/03/25/drilling-of-well-56-32/)](utah_forge_56_32.png){fig-align='center'}

1. [1 pt] Read in the data. 

```{r, indent = "   ", message = F, warning = F}

```

2. [3 pt] Plot the time series and describe the time plot in terms of the trend and seasonality.

```{r, indent = "   ", fig.dim = c(6, 4)}

```

   :::{.callout-warning icon=false appearance="simple"}
   
   :::

3. [1 pt] Hypothesize a reasonable seasonal frequency for this data set and justify your choice. 

   :::{.callout-warning icon=false appearance="simple"}
   
   :::

4. [2 pt] Create two time series objects: 1) describing the total mud volume until (but not including) 03:50:00 (called `utah_train_ts`) and 2) describing the total mud volume from 03:50:00 onward (called `utah_test_ts`).

```{r, indent = "   ", message = F, warning = F}

```

5. [1 pt] Create boxplots of the total volume of mud by each element of the seasonal cycle for the `utah_train_ts` time series.

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

   
6. [2 pt] Plot a decomposition of the `utah_train_ts` time series and component on what you see in terms of trend and seasonality. Does the decomposed seasonal trend represent meaningful seasonal variation? Explain your answer. 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

   :::{.callout-warning icon=false appearance="simple"}

   :::
   
7. [2 pt] Create an ACF plot of the residual error series from the decomposition and comment on what the plot suggests about serial correlation in the residual error series.

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

   :::{.callout-warning icon=false appearance="simple"}

   :::   

8. [2 pt] Could you use the decomposition model to forecast the total mud volume values in the `utah_test_ts` series? Why or why not?

   :::{.callout-warning icon=false appearance="simple"}
  
   :::   

9. [5 pt] Fit an additive Holt-Winters model to the `utah_train_ts` series, allowing R to estimate the model parameters. Use that model to forecast the total volume of mud in the `utah_test_ts` series, and plot the observed, estimated, and forecasted series on a single plot, including 95% prediction intervals for the forecast. 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

10. [2 pt] Comment on the quality of the forecast, discussing both the forecasted total mud volume, and the uncertainty in that forecast.

    :::{.callout-warning icon=false appearance="simple"}
  
    :::  

11. [2 pt] Propose a way to assess the quality of the forecast that incorporates **both** the forecasted volume (the point estimate) **and** the uncertainty in that forecasted volume (the interval estimate). **Note:** it is okay if your loss function results in two numbers. 

    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 
    
12. [1 pt] Implement your proposed loss function on the Holt-Winters model estimated by R and print out the value of that loss function.
    
```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```

13. [5 pt] Find values of the Holt-Winters model parameters that result in a better forecast, according to the criterion you proposed. Plot the observed, estimated, and forecasted volumes from this model as before, and comment on why the new fit is better. 

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```

    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 
    
14. [2 pt] Fit an AR(1) model to the `utah_train_ts` series and allow R to estimate the smoothing parameter. Write out the estimated AR model.

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```

    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 

15. [2 pt] Does the estimated AR model represent a stationary process? Why or why not?

    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 
    
    
16. [2 pt] Create an acf plot of the residuals from the AR model and comment on what this plot means in terms of residual autocorrelation. 

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```

    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 
    

17. [5 pt] Use the AR model to forecast the `utah_test_ts` series and plot the observed, estimated, and forecasted series on a single plot. Comment on the quality of the forecast.

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```
    
    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 
    
18. [2 pt] You should have come to the conclusion that neither the AR(1) model nor the Holt-Winters model provided a very good forecast for these data. Comment on why that might be the case, referencing both the structure of the observed series, and the length of the period over which we are attempting to forecast.     
    
    :::{.callout-warning icon=false appearance="simple"}
  
    ::: 
    
\newpage

**Question 2 [12 pt]**

For each of the following AR models, write the model in terms of the backshift operator and determine whether the model is stationary.  

1. [3 pt] $x_t = .25x_{t-1} + w_t$

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 
   
```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```
   
2. [3 pt] $x_t = -1.1x_{t-1} + w_t$

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 
   
```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```


3. [3 pt] $x_t = -1.1x_{t-1} + .25x_{t-2} + w_t$

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

4. [3 pt] $x_t = 1.2x_{t-1} - .3x_{t-2} + w_t$

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

\newpage

**Question 3 [25 pt]**

We discussed in class how useful simulation can be for understanding time series concepts. In this question, we will use simulation to better understand how to construct time series using AR processes. 

1. [3 pt] Use simulation to generate 24 years of monthly data from an AR(2) process with $\alpha_1 = .7$, $\alpha_2 = -.3$, and $\sigma = 5$. Create a time series object called `error_ts` and plot that series. 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

2. [2 pt] Create an autocorrelation function plot of `error_ts` and comment on what the plot suggests about the residual correlation in the series. 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

3. [2 pt] Create a partial autocorrelation function plot of `error_ts` and comment on what the plot suggests about the residual correlation in the series. 

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

4. [1 pt] Is the series constructed in question 1 capable of modeling time series with trends or seasonal components? Why or why not?

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 
   
5. [2 pt] A time series may be constructed by adding a serially correlated error series (such as `error_ts`) to a model that describes the mean function over time. Regression is often used to model the mean function in time. For each time point in the 24 years simulated in the `error_ts` series ($t = 1, 2, \dots, 288$), compute $.5t - .001t^2 + 5\sin\left(\frac{2\pi t}{12}\right) - 5\cos\left(\frac{2 \pi t}{12}\right)$ and plot the resulting function with respect to time. 
   
```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```
   
6. [1 pt] The function plotted in the previous question is a smooth curve with no noise, rather than the noisy series we see with observed data. Why is that?

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

7. [2 pt] Add the function created in question 5 to `error_ts`. **Note:** You will likely have to convert `error_ts` to a vector using `c()` before adding the two series, then recreate a time series object after summing. Call this new series `sim_ts`. Plot the resulting series.

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4)}

```


8. [1 pt] Comment on the plot in the previous question. How does it compare to some of the series we have seen in this class so far?

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

9. [1 pt] The code below fits a regression model with $t$, $t^2$, $\sin(t/4)$, and $\cos(t/4)$ as explanatory variables for the `sim_ts` time series and prints off the coefficient estimates. How do these values compare to the values used to generate the time series? 

   :::{.callout-important title="STOP"}
   After you have created the `sim_ts` object, you will have to change the `eval = F` chunk option to `eval = T`.
   :::

```{r, indent = "   ", message = F, warning = F, fig.dim = c(6, 4), eval = F}
reg_tbl <- tibble(
  y = c(sim_ts)
) %>%
  mutate(
    t = 1:n(),
    sint = sin(2*pi*t/12),
    cost = cos(2*pi*t/12)
  )

sim_reg <- lm(
  y ~ t + I(t^2) + sint + cost,
  data = reg_tbl
)
sim_reg$coefficients[-1]
```

   :::{.callout-warning icon=false appearance="simple"}
   
   ::: 

10. [2 pt] The code below calculates the residuals from the fitted regression model, which represents our residual error series. Create acf and pacf plots of the residual error series, and compare these plots to the acf and pacf plots created for the `error_ts` series. 

    :::{.callout-important title="STOP"}
    You will have to change the `eval = F` chunk option to `eval = T`.
    :::

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 7), eval = F}
res <- resid(sim_reg)
par(mfrow = c(2, 1))

par(mfrow = c(1,1))
```

    :::{.callout-warning icon=false appearance="simple"}
    
    ::: 
    
11. [2 pt] Fit an AR(2) model to the residual error series, called `ar_fit`. Print the estimated of $\alpha_1$ and $\alpha_2$ and comment on how these estimates relate to the values used to generate the error series in question 1. 

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```

    :::{.callout-warning icon=false appearance="simple"}
    
    ::: 

For the remainder of this question, assume we are trying to forecast two more years of the simulated series. To do so, we must first generate two more years of the series, corresponding to $t = 289, \dots, 312$.

12. [2 pt] The code below generates two more years of the `error_ts` series and creates a new series, `full_error_ts`, that includes all 26 years of data. Describe what the first four lines of code are doing and why they are necessary to maintain the AR(2) process. 

    :::{.callout-important title="STOP"}
    You will have to change the `eval = F` chunk option to `eval = T` after completing the previous questions.
    :::

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4), eval = F}
# simulate two more years
x <- w <- rnorm(24, 0, 5)
x[1] <- .7*error_ts[278] - .3*error_ts[277] + rnorm(1, 0, 5)
x[2] <- .7*x[1] - .3*error_ts[278] + rnorm(1, 0, 5)
for(t in 3:24) x[t] <- .7*x[t-1] - .3*x[t-2] + w[t]

# combine series and plot
full_error_ts <- ts(
  c(error_ts, x),
  start = c(1, 1),
  freq = 12
)
plot(full_error_ts)
abline(v = 25, lty = 2)
```

    :::{.callout-warning icon=false appearance="simple"}
    
    ::: 

13. [2 pt] Generate two more years ($t = 289, 290, \dots, 312$) from the mean function defined in question 5. Create a new series, `full_ts`, that combines the mean function from $t=1$ to $t=312$ with the `full_error_ts` series, and plot that series. 

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4)}

```

14. [2 pt] The code below obtains predictions for $t = 289, \dots, 312$ from the fitted regression model and the AR process fit to the residual error series, and combines them to forecast the series for $t = 289, \dots, 312$. Comment on the quality of the forecast.


    :::{.callout-important title="STOP"}
    You will have to change the `eval = F` chunk option to `eval = T` after completing the previous questions.
    :::

```{r, indent = "    ", message = F, warning = F, fig.dim = c(6, 4), eval = F}
# obtain forecast of mean function from the regression model
pred_tbl <- tibble(
  t = 289:312
) %>%
  mutate(sint = sin(2*pi*t/12), cost = cos(2*pi*t/12))
pred <- predict(sim_reg, newdata = pred_tbl, se.fit = T)

# predict from ar_fit
pred_error <- predict(ar_fit, n.ahead = 24, prediction.interval = T)

# combine them
forecast_series <- ts(
   c(pred$fit + pred_error$pred),
   start = c(25, 1),
   freq = 12
)
forecast_series_lwr <- ts(
  c(forecast_series) - 2*sqrt(pred$se.fit^2 + c(pred_error$se^2)),
   start = c(25, 1),
   freq = 12
)
forecast_series_upr <- ts(
  c(forecast_series) + 2*sqrt(pred$se.fit^2 + c(pred_error$se^2)),
   start = c(25, 1),
   freq = 12
)

# plot
plot(full_ts)
abline(v = 25, lty = 2)
lines(forecast_series, lty = 1, col = 2)
lines(forecast_series_lwr, lty = 2, col = 2)
lines(forecast_series_upr, lty = 2, col = 2)
``` 

    :::{.callout-warning icon=false appearance="simple"}
    
    ::: 
