---
title: "Homework 3"
format:
  pdf:
    fontsize: 12pt
    geometry:
      - inner=1.5cm
      - outer=1.5cm
      - top=2.5cm
      - bottom=2.5cm
    include-in-header:
      - text: |
          \addtokomafont{disposition}{\rmfamily}
          \usepackage{fancyhdr, lastpage, framed, caption, xcolor}
          \captionsetup[figure]{labelformat=empty}
          \pagestyle{fancyplain}
          \fancyhf{}
          \lhead{\fancyplain{}{STAT 0219: Time Series Analysis}}
          \rhead{\fancyplain{}{Stratton - HW2}}
          \fancyfoot[RO, LE] {page \thepage\ of \pageref{LastPage}}
          \thispagestyle{plain}
editor: source
---
<!-- Note: the YAML header is everything above this line. -->
\vspace{-3in}
\textbf{Name:} Your name here  
\textbf{Due:} 2024/09/30
\vspace{1.5in}

Be sure to submit **both** the .pdf and .qmd file to Canvas by Monday, September 30th at 11:59 pm. 

```{r, include = F}
rm(list = ls())
library(tidyverse)
```

0. [1 pt] With whom did you work on this assignment?

   :::{.callout-warning icon=false appearance="simple"}
   
   :::

1. [11 pt] We will focus on a [data set](https://www.kaggle.com/datasets/neuromusic/avocado-prices) describing weekly avocado sales volume and price in the United States between 2015 and 2018 for this question. 

   a) [2 pt] Read the data in (naming it `avocado`), filter to sales of conventional avocados in Las Vegas, and create a new column, called `volume1000`, that represents the total volume of sales in 1000s. 

```{r, indent = "      ", message = F, warning = F}

```


   b) [3 pt] Create a `ts` object with the `volume1000` vector (called `avo_ts`), create an additive decomposition of that time series (called `avo_decomp`), and plot that decomposition.
   
      :::{.callout-tip}
      Be sure to pay attention to how the data set is arranged with respect to date. 
      :::
   
```{r, indent = "      ", fig.dim = c(6, 5)}

```


   c) [1 pt] Describe the time series decomposition in terms of its trend and seasonal components. 

      :::{.callout-warning icon=false appearance="simple"}

      :::
      
      
   d) [1 pt] You should notice that the trend and random components are significantly (in the english way, not the statistics way) shorter than the other two series. Why is this?
   
      :::{.callout-warning icon=false appearance="simple"}

      :::

   e) [1 pt] You should see a rather large spike in avocado sales during the beginning of each year in the decomposition. Hypothesize an explanation for this spike.
   
      :::{.callout-warning icon=false appearance="simple"}

      :::
      
   f) [1 pt] The code below attempts to create a data frame of the random component over time, but R kicks back an error. Investigate the source of this error, and describe what is causing the problem. 
   
```{r, indent = "      ", eval = F}
res_tbl <- tibble(
  dt = avocado$Date,
  res = avo_decomp$random
)
```

      :::{.callout-warning icon=false appearance="simple"}

      :::

```{r, indent = "      "}

```

   g) [2 pt] We are going to ignore the problem for now ( :] ). Provide a correlogram of the residual error series and comment on whether there appears to be leftover serial correlation. 
   
```{r, indent = "      ", fig.dim = c(6, 4)}

```

      :::{.callout-warning icon=false appearance="simple"}

      :::

2. [1 pt] The code below creates a time series out of the same seven values on repeat, and generates a correlogram for that series. Describe what is happening in the correlogram at time lags that are a multiple of seven. Why should that make sense?

```{r, indent = "   ", fig.dim = c(6, 4)}
ex_ts <- ts(
  rep(-3:3, 8),
  start = c(1,1),
  end = c(8, 7),
  freq = 7
)
acf(ex_ts)
```

   :::{.callout-warning icon=false appearance="simple"}

   :::

3. [6 pt] The rest of the questions on this assignment are fairly math-heavy, but in order to understand time series analysis, it is essential to understand the concepts of stationarity, covariance, and correlation. If the following questions feel hard, *that is okay* - it does not mean you cannot learn time series. I encourage you to ask for help! 
   
   **Note:** When I use the word *prove* below, I do not mean in a rigorous mathematical sense (though, by all means give it a shot!). I only ask that you provide logic and sound mathematical reasoning (meaning I should see an equation or two in each response).

   a) [3 pt] *Prove* that a model cannot be second-order stationary if the model is not stationary in the mean. 
   
      :::{.callout-tip title="Hint"}
      What do we know about the autocovariance function, $\gamma(k)$ for a second-order stationary process?
      :::
      
      :::{.callout-warning icon=false appearance="simple"}

      :::

   b) [3 pt] Recall that the population autocorrelation function is given by $\rho_k = \frac{\gamma_k}{\sigma^2}$. *Prove* that $\rho_0$ must always equal 1. 
   
      :::{.callout-warning icon=false appearance="simple"}

      :::

4. [4 pt] (This question is admittedly a nasty piece of work, but valuable to understand). Recall that the sample autocorrelation function, $r_k$, is defined as
   $$
   r_k = \frac{c_k}{c_0}
   $$
   but the population autocorrelation function, $\rho_k$ is defined as 
   $$
   \rho_k = \frac{\gamma_k}{\sigma^2}.
   $$
   Why do we define $r_k = \frac{c_k}{c_0}$, rather than $r_k = \frac{c_k}{s^2}$, which is arguably more natural? Support your answer with something proof-ish.
   
   :::{.callout-warning icon=false appearance="simple"}
   
   :::
   


