---
title: "Day 3 - Decomposition of time series"
format:
  pdf:
    fontsize: 12pt
    geometry:
      - inner=1.5cm
      - outer=1.5cm
      - top=2.5cm
      - bottom=2.5cm
    include-in-header:
      - text: |
          \addtokomafont{disposition}{\rmfamily}
          \usepackage{fancyhdr, lastpage, framed, caption, xcolor}
          \captionsetup[figure]{labelformat=empty}
          \pagestyle{fancyplain}
          \fancyhf{}
          \lhead{\fancyplain{}{STAT 0219: Time Series Analysis}}
          \rhead{\fancyplain{}{Stratton - Day 3}}
          \fancyfoot[RO, LE] {page \thepage\ of \pageref{LastPage}}
          \thispagestyle{plain}
editor: source
---
\vspace{-1in}

## Introduction

The purpose of today's lecture is to understand how to decompose a time series into its constituent components in R. To guide our exploration, we will return to the Vermont temperatures and Pan Am data sets.  

```{r, echo = T, warning = F, message = F}
# packages
library(tidyverse)

# load data and rename
## Pan Am
data(AirPassengers)
ap <- AirPassengers

## vt temps
vt_temps <- readr::read_csv("vt_temps.csv")
```

***

\newpage

## Review: creating `ts` objects

Create a `ts` object, called `vt_ts`, for the monthly temperatures in Vermont that spans from 1970/06/01 to 2013/04/01. Plot the time series. 

```{r, fig.dim = c(7, 3.75)}
# construct a ts object from montly vt_temps spanning  06/1850 - 06/2013
vt_temps_clean <- vt_temps %>%
  dplyr::select(dt, avg_temp = AverageTemperature) %>%
  mutate(
    dt = lubridate::ymd(dt)
  ) %>%
  filter(
    dt >= lubridate::ymd("1970-06-01"), 
    dt <= lubridate::ymd("2013-04-01")
  )

vt_ts <- ts(
  vt_temps_clean$avg_temp,
  start = c(1970, 6),
  end = c(2013, 4),
  frequency = 12
)

plot(vt_ts, ylab = "Average monthly temperature (C)")
```

```{r, eval = F, echo = T}
# alternative using window()
vt_ts_long <- ts(
  vt_temps$AverageTemperature,
  start = c(1850, 1),
  end = c(2013, 9),
  freq = 12
)

vt_ts <- window(
  vt_ts_long, 
  start = c(1970, 6),
  end = c(2013, 4)
)
```

## Introducing definitions and notation 

:::{.callout-note title="Random variables"}
A ____________________, usually written $X$, is a variable whose possible values are the numerical outcomes of ___________ phenomenon. There are two types of random variables, ______________ and ____________.

\vspace{.125in}

Formally, a random variable is a mapping from a sample space $S$ to the real numbers.
:::

\vspace{.125in}

**Discrete random variables:** \hspace{1.5in} **Continuous random variables:**

\vspace{2in}


:::{.callout-note title="Time series notation"}
A ____________________ of length $n$ is a sequence of ________________, which we denote $\{X_t : t = 1, \dots, n\} = \{X_1, X_2, \dots, X_n\}$. When referring to an observed time series, we use lowercase letters, $\{x_t : t = 1, \dots, n\} = \{x_1, x_2, \dots, x_n\}$. If the length of the series $n$ does not need to be specified, we will often use the abbreviated notation $\{x_t\}$. 
:::

- $\bar{x} = \frac{\sum x_i}{n}$ 

\vspace{.75in}

- $\hat{x}$

\vspace{.75in}

- $\hat{x}_{t+k|t}$

\vspace{.75in}

## Our first time series model

:::{.callout-note title="Decomposition models"}
An _______________________________ is a simple model for a time series that estimates the ____________, _______________________, and ____________. 

$$
x_t = m_t + s_t + z_t
$$

A ____________________________________ allows for the seasonal effect to increase as the trend increase. 

$$
x_t = m_t \cdot s_t \cdot z_t
$$

If the time series is strictly positive, it may be easier to fit an additive model on the log scale than a multiplicative model on the original scale. 

$$
\log(x_t) = m_t + s_t + z_t
$$
:::

\newpage

## Estimating $m_t$, $s_t$, and $z_t$

How can we obtain an estimate of the trend effect?

\vspace{1.5in}


:::{.callout-note title="Centered moving average"}
For time series with a period of 12 (i.e. monthly data), the ______________________ at time $t$ is given by
$$
\hat{m}_t = \frac{\frac{1}{2}x_{t-6}+x_{t-5}+\dots+x_{t-1}+x_t+x_{t+1}+\dots+x_{t+5}+\frac{1}{2}x_{t+6}}{12}
$$
where $t=7, \dots, n-6$
:::

\vspace{.125in}

How can we obtain an estimate of the seasonal effect at each time $t$? How can we obtain an estimate of the overall seasonal effect associated with each month?

\vspace{1.5in}

:::{.callout-note title="Seasonal effects"}
For an additive time series with a monthly frequency, the seasonal effect at time $t$ is estimated by 
$$
\hat{s}_t = x_t - \hat{m}_t
$$
We can obtain a single estimate of the monthly effect by averaging the effect of each month. 
$$
\bar{s}_t = \frac{\sum s_t}{T-1}
$$
where $T$ denotes the number of years. Often times, the estimated seasonal effect is **centered** after calculation - more on this on Wednesday. If a time series is multiplicative, the seasonal effect is instead estimated by 
$$
\hat{s}_t = \frac{x_t}{\hat{m}_t}
$$
:::

\newpage

How can we obtain an estimate of $z_t$?

\vspace{1in}

:::{.callout-note title="Residual error series"}
The _________________________, also called _____________, is the raw time series adjusted for the trend and seasonal effects. On average, this series should have a mean of ____________. For an additive decomposition model, the residual error series is
$$
\hat{z}_t = x_t - \hat{m}_t - \bar{s}_t
$$
For a multiplicative decomposition model, the residual error series is
$$
\hat{z}_t = \frac{x_t}{\hat{m}_t \cdot \bar{s}_t}
$$
:::

\newpage

## Decomposition in R

:::{.callout-note}
The `decompose` function may be used in R to obtain a decomposition of a time series object. 
:::

```{r, fig.dim = c(7, 7)}
plot(decompose(ap))
```

\newpage

```{r, fig.dim = c(7, 7)}
plot(decompose(ap, type = "multiplicative"))
```

\newpage

```{r, fig.dim = c(7, 7)}
plot(decompose(log(ap)))
```


